Measurement of ($\pi^-$-Ar) and ($K^+$-Ar) total hadronic cross sections at the LArIAT experiment

### 2018-05-30 Chapter-4/4-Methodology.tex
s

\subsection{TPC Selection: Shower Removal}\label{ch:electrons}
In the case of the ($\pi^-$,Ar) cross section, the resolution of  beamline mass spectrometer is not sufficient to select a beam of pure pions. In fact,  muons which are close in mass to the pions and relativistic electrons survive the selection on the beamline mass.  It is important to notice that the composition of the negative polarity beam is mostly pions, as will be discussed in section \ref{ch:beamlineComposition}.
Still, we devise a selection on the TPC information to mitigate the presence of electrons in the sample used for the pion cross section. The selection relies on the different topologies of a pion and an electron event when propagating in liquid argon: while the former will trace a track inside the TPC active volume, the latter will tend to ``shower", i.e. interact with the medium, producing bremsstrahlung photons which pair convert into several short tracks. In order to remove the shower topology, we create a region of interest (ROI) around the TPC track corresponding to the beamline particle. We look for short tracks contained in the ROI, as depicted in figure \ref{fig:showerFilt}:  if more then 5 tracks shorter than 10 cm are in the ROI, we reject the event. Line 7 in in Table \ref{tab:beamlineDataSelection} shows the number of events surviving this selection.

@@

\subsection{Not-so-Thin Target: Slicing the Liquid Argon Volume}\label{ch:XSRaw}
The interaction length of pions and kaons in liquid argon is expected to be of the order of 50 cm for pions and 100 cm for kaons. Thus, the LArIAT TPC, with its 90 cm of length, is not a thin target. However, the fine-grained tracking of the LArIAT LArTPC allows us to treat the argon volume as a sequence of many adjacent thin targets.

As described in Chapter \ref{sec:experimentDescription}, LArIAT induction and collection planes consist of 240 wires each at 4 mm spacing. The wires are oriented at +/- $60^{\circ}$ from the vertical direction, while the beam direction is oriented 3 degrees off the $z$ axis in the $XZ$ plane.  The collection wires collect signals proportional to the energy deposited by the hadron along its path in a  $\delta${\emph{X}} = 4 mm/(sin($60^{\circ}$)cos($3^{\circ}$)) $\approx$ 4.7~mm slab of liquid argon. Thus, one can think to slice the TPC into many thin targets of $\delta${\emph{X}} = 4.7~mm thickness along the direction of the incident particle, making a measurement at each wire along the path.

>>>>

\subsection{TPC Selection: Shower Removal}\label{ch:electrons}
In the case of the ($\pi^-$,Ar) cross section, the resolution of  beamline mass spectrometer is not sufficient to select a beam of pure pions. In fact,  muons which are close in mass to the pions and relativistic electrons survive the selection on the beamline mass.  It is important to notice that the composition of the negative polarity beam is mostly pions, as will be discussed in section \ref{ch:beamlineComposition}.
Still, we devise a selection on the TPC information to mitigate the presence of electrons in the sample used for the pion cross section. The selection relies on the different topologies of a pion and an electron event when propagating in liquid argon: while the former will trace a track inside the TPC active volume, the latter will tend to ``shower", i.e. interact with the medium, producing bremsstrahlung photons which pair convert into several short tracks. In order to remove the shower topology, we create a region of interest (ROI) around the TPC track corresponding to the beamline particle. We look for short tracks contained in the ROI, as depicted in figure \ref{fig:showerFilt}:  if more then 5 tracks shorter than 10 cm are in the ROI, we reject the event. Line 7 in  Table \ref{tab:beamlineDataSelection} shows the number of events surviving this selection; that table also shows that this selection is applied after the wire chamber to TPC match (discussed in the next section). The WC2TPC match already lower the presence of electrons in the sample, which is further reduced by the shower filter.

@@

\subsection{Not-so-Thin Target: Slicing the Liquid Argon Volume}\label{ch:XSRaw}
The interaction length of pions and kaons in liquid argon is expected to be of the order of 50 cm for pions and 100 cm for kaons. Thus, the LArIAT TPC, with its 90 cm of length, is not a thin target. However, the granularity of the LArIAT LArTPC detector allows us to treat the argon volume as a sequence of many adjacent thin targets.

As described in Chapter \ref{sec:experimentDescription}, LArIAT induction and collection planes consist of 240 wires each at 4 mm spacing. The wires are oriented at +/- $60^{\circ}$ from the vertical direction, while the beam direction is oriented 3 degrees off the $z$ axis in the $XZ$ plane.  The collection wires collect signals proportional to the energy deposited by the hadron along its path in a  $\delta${\emph{X}} = 4 mm/(sin($60^{\circ}$)cos($3^{\circ}$)) $\approx$ 4.7~mm slab of liquid argon. Thus, one can think to slice the TPC into many thin targets of $\delta${\emph{X}} = 4.7~mm thickness along the direction of the incident particle, making a measurement at each wire along the path, as sketched in Figure \ref{fig:TPCGran}.

\begin{figure}
  \centering  
\includegraphics[width=0.8\textwidth]{Chapter-4/Images/LArSlice.png}
\caption{Representation of sliced LAr Volume.}
\label{fig:TPCGran}
\end{figure}

### 2018-05-30 Chapter-6/6-PionCrossSection.tex
s

\clearpage
\section{Results}\label{ch:FinalPion}
Figure \ref{fig:FinalXSPion} show the measurement of the ($\pi^-$-Ar) total hadronic cross section for  scattering angles greater than 5$^\circ$, as the result of the background subtraction and reconstruction effects correction to the raw cross section. The top left plot is the measurement obtained on the 60A data, statistical uncertainty in black and systematic uncertainty in red. The top right plot is the measurement obtained on the 100A data, statistical uncertainty in black and systematic uncertainty in blue. The bottom plot shows the two measurements overlaid. In all three plot, the Geant4 prediction for the total hadronic cross section for angle scattering greater than 5$^\circ$ is displayed in green.

@@

@@

>>>>

\clearpage
\section{Results}\label{ch:FinalPion}
Figure \ref{fig:FinalXSPion} shows the measurement of the ($\pi^-$-Ar) total hadronic cross section for  scattering angles greater than 5$^\circ$, as the result of the background subtraction and reconstruction effects correction to the raw cross section. The top left plot is the measurement obtained on the 60A data, statistical and systematic uncertainty shown in red. The top right plot is the measurement obtained on the 100A data, statistical uncertainty and systematic uncertainty shown in blue. The bottom plot shows the two measurements overlaid. In all three plot, the Geant4 prediction for the total hadronic cross section for angle scattering greater than 5$^\circ$ is displayed in green.

@@

The cross section measurements in the two datasets agree within the systematic uncertainty, even if the cross section measured in the 60A data is lower than the one measured in the 100A data in all overlapping bins except for the 200-250 MeV bin.

@@

The top panel of Figure \ref{fig:FinalXSResPion} shows the cross section obtained combining the two datasets. To combine the datasets, we merge the corrected interacting histogram for the 60A data with the interacting histogram for the 100A data; we apply the same merging with the incident histograms. We then calculate the cross section central point and statistical uncertainty using the merged interacting and incident histograms. The systematic uncertainty is calculated as the sum in quadrature of the original systematic uncertainties on the cross section measurements on the two separated data sets.

The bottom panel of figure  \ref{fig:FinalXSResPion} shows the relative difference between the measured cross section and the Geant4 predicted cross section.
A couple of considerations are in order here. On one hand, we note the general agreement between the measured and predicted cross sections. We expected the pion cross section in argon to be well modeled in Geant4, given the vast set of measurements available for pion interaction -- albeit not on argon. Thus, this agreement augments our confidence in the experimental methodology used.
On the other hand, this agreement is not perfect.  As a matter of fact, its imperfections are rather suggestive: we see a slight shift of the delta resonance at higher energies and a small shape difference, with a steeper drop between the delta region and the flatter part at energies greater than 400 MeV. 
Since we see room for improvement of the systematic uncertainties, we look forward to see if this difference in shape will be confirmed by further analyses.

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{Chapter-6/Images/TheRealMoneyPlot.pdf}
\caption{ \emph{Top:} ($\pi^-$-Ar) total hadronic cross section for  scattering angles greater than 5$^\circ$ measured in the combined sample, statistical uncertainty and systematic uncertainty in black. The Geant4 prediction for the total hadronic cross section for angle scattering greater than 5$^\circ$ is displayed in green. \emph{Bottom:} Relative difference between the measured cross section and the Geant4 prediction. }
\label{fig:FinalXSResPion}
\end{figure}

### 2018-05-30 thesis_main.tex
s

\linenumbers

@@

% Need to define title before the abstract.
\title{Measurement of ($\pi^-$-Ar) and ($K^+$-Ar) total hadronic cross sections in the LArIAT experiment}
\author{Elena Gramellini}
\advisor{Bonnie T. Fleming}
\date{Date you'll receive your degree} % usually not \today.

@@

The Liquid Argon Time Projection Chamber (LArTPC) represents one of the
most advanced experimental technologies for physics at the Intensity
Frontier due to its full 3D-imaging, excellent particle identification
and precise calorimetric energy reconstruction. By deploying a LArTPC in
a dedicated calibration test beam line at Fermilab, LArIAT (Liquid
Argon In A Testbeam) aims to experimentally calibrate this technology
in a controlled environment and to provide physics results key to the
neutrino oscillation physics and proton decay searches of the Short
Baseline Neutrino and Long Baseline Neutrino programs.

@@

This thesis presents two world's first measurements: the measurement of ($\pi^-$-Ar) total hadronic cross section in the 100-1050 MeV kinetic energy range and the measurement of the ($K^+$-Ar) total hadronic cross section in the 100-650 MeV kinetic energy range. The analyses devised for these measurements use both the core elements of LArIAT: beamline and TPC. The first step in each analysis is the development of an event selection based on  beamline and TPC information geared towards the identification of the hadron of interest. We then proceed to match the beamline candidate to a suitable TPC track. Finally,  we apply the ``thin slice method" technique and measure the cross section, correcting for background and detector effects. The thin slice technique is a new method to measure hadron-argon cross sections possible only thanks to the combination of the tracking and calorimetry capability of the LArTPC technology. 
Albeit never on argon, the hadronic cross section of pions has been measured before on several different elements in thin target experiments, leading to solid predictions in the argon case. Through the use of a different technique, our measurement of the  ($\pi^-$-Ar) total hadronic cross section is in general agreement with the prediction by the Geant4 Bertini Cascade model which are based on data from thin target experiments. On the contrary, cross section measurements for kaons are extremely rare, thus more difficult to model. Not surprisingly, our measurement of the  ($K^+$-Ar) total hadronic cross section is mostly in disagreement with the Geant4 prediction.

>>>>

%\linenumbers

@@

% Need to define title before the abstract.
\title{Measurement of ($\pi^-$-Ar) and ($K^+$-Ar) total hadronic cross sections at the LArIAT experiment}
\author{Elena Gramellini}
\advisor{Bonnie T. Fleming}
\date{Date you'll receive your degree} % usually not \today.

@@

The Liquid Argon Time Projection Chamber (LArTPC) represents one of the
most advanced experimental technologies for physics at the Intensity
Frontier due to its full 3D-imaging, excellent particle identification
and precise calorimetric energy reconstruction. By deploying a LArTPC in
a dedicated calibration test beam line at Fermilab, LArIAT (Liquid
Argon In A Testbeam) aims to experimentally calibrate this technology
in a controlled environment and to provide physics results key to the
neutrino oscillation physics and proton decay searches of the Short
Baseline Neutrino (SBND, MicroBooNE, ICARUS) and Long Baseline Neutrino programs (DUNE).

@@

This thesis presents two world's first measurements: the measurement of ($\pi^-$-Ar) total hadronic cross section in the 100-1050 MeV kinetic energy range and the measurement of the ($K^+$-Ar) total hadronic cross section in the 100-650 MeV kinetic energy range. The analyses devised for these measurements use both the core elements of LArIAT: beamline and TPC. The first step in each analysis is the development of an event selection based on  beamline and TPC information geared towards the identification of the hadron of interest. We then proceed to match the beamline candidate to a suitable TPC track. Finally,  we apply the ``thin slice method" technique and measure the cross section, correcting for background and detector effects. The thin slice technique is a new method to measure hadron-argon cross sections possible only due to the combination of the tracking and calorimetry capability of the LArTPC technology. 
Albeit never on argon, the hadronic cross section of pions has been measured before on several different elements in thin target experiments, leading to solid predictions for measurements on argon. Through the use of a different technique, our measurement of the  ($\pi^-$-Ar) total hadronic cross section is in general agreement with the prediction by the Geant4 Bertini Cascade model which are based on data from thin target experiments. On the contrary, cross section measurements for kaons are extremely rare, thus more difficult to model. Our measurement of the  ($K^+$-Ar) total hadronic cross section is mostly in disagreement with the Geant4 prediction and provides new experimental data to properly tune the Geant4 models.
